---
title: "PA1_resub20180901"
author: "Harold Pollack"
date: "9/2/2018"
output: html_document
---

This R markdown file explores an activity dataset. It compares activity on the weekdays and weekends. It imputes values, and gives an opportunity to use GitHub. This is a resubmission.

First I load libraries

```{r setup, include=FALSE}
library("dplyr", lib.loc="/Library/Frameworks/R.framework/Versions/3.4/Resources/library")
library("forecast", lib.loc="/Library/Frameworks/R.framework/Versions/3.4/Resources/library")
library("knitr", lib.loc="/Library/Frameworks/R.framework/Versions/3.4/Resources/library")
library("ggplot2", lib.loc="/Library/Frameworks/R.framework/Versions/3.4/Resources/library")
library("xts", lib.loc="/Library/Frameworks/R.framework/Versions/3.4/Resources/library")
library(lattice)
library("gridExtra", lib.loc="/Library/Frameworks/R.framework/Versions/3.4/Resources/library")
library("ggpubr", lib.loc="/Library/Frameworks/R.framework/Versions/3.4/Resources/library")

knitr::opts_chunk$set(echo = TRUE)
```

Now I read read and processe activity.csv to read the activity dataset. I also construct a logged version of steps, which I will use below.

```{r read dataset, cache=TRUE}
activity <- read.csv("/Users/haroldpollack/Documents/coursera_datascience3/activity.csv")
activity$lsteps<-log(1+activity$steps)
activity$day <-weekdays(as.Date(activity$date))
table(activity$day)
str(activity)
summary(activity)
```

```{R mean and median, cache=TRUE}
step_mean <- mean(activity$steps, na.rm = TRUE)
step_median <- median(activity$steps, na.rm = TRUE)
```
Now the mean number of steps is `r step_mean`.

The median number of steps is `r step_median`.

## Now let's show some graphs of the step distribution


```{r Histograms on the level and log scale, echo=FALSE, cache=TRUE}
qplot(activity$steps, 
      geom="histogram",
      binwidth=50,  
      main="Histogram of Steps over the entire dataset", 
      fill=I("blue"), 
      xlab = "Steps"
      ) 
## log of steps
# hist(activity$lsteps, 
#      breaks = 10,  
#      main="Histogram of Steps shown on log scale", 
#      xlab = "log(1+Steps)"
#      ) 
```

## Find which days are weekends

```{R weekends, cache=TRUE}
charday <- as.character(activity$day)
table(charday)
dSunday<-as.numeric((activity$day=="Sunday"))
dMonday<-as.numeric((activity$day=="Monday"))
dTuesday<-as.numeric((activity$day=="Tuesday"))
dWednesday<-as.numeric((activity$day=="Wednesday"))
dThursday<-as.numeric((activity$day=="Thursday"))
dFriday<-as.numeric((activity$day=="Friday"))
dSaturday<-as.numeric((activity$day=="Saturday"))
dweekend<-dSunday+dSaturday
activity$dweekend <- dweekend
table(dweekend)
```

Now we have to handle missing values. I replaced with nonnegative predicted values based on a simple regression with day dummies and interval dummies. I then forced the values to be non-negative. A slightly better model would employ censored regression or something like that.


```{R missing values and aggregation, cache=TRUE}
#
#  regression model with non-interacted day and interval dummies
#
model_steps <- lm(steps ~ as.factor(day)+as.factor(interval), data=activity)
# summary(model_steps)
activity$missing_steps <- as.numeric(is.na(activity$steps))
table(activity$missing_steps)
n_missing <- sum(activity$missing_steps)
# n_missing
activity$pred_steps=predict(model_steps,activity)
summary(activity$pred_steps)
str(activity$pred_steps)
activity$imputed_steps<-activity$steps
summary(activity$imputed_steps)
str(activity$imputed_steps)
#
# loop through each observation
#
for(i in 1:nrow(activity))
{
  activity$imputed_steps[i]<- activity$steps[i]
  if(activity$missing_steps[i] == 1)
  {
    # 
    #    Note that I want predicted values to be nonnegative. Which I do below.
    #
    activity$imputed_steps[i] <- max(0,activity$pred_steps[i])
  }
}
summary(activity$imputed_steps_missing)
str(activity$imputed_steps_missing)

summary(activity$steps)
summary(activity$imputed_steps)
summary(activity$missing_steps)
str(activity$steps)
str(activity$imputed_steps)
str(activity$missing_steps)
imputed_steps <- activity$imputed_steps
activity_date <- activity$date
interval <-activity$interval
```

Note that there are `r n_missing' missing steps. 
Now deal with weekend stuff in a clunky way by defining two variables.

```{R now deal with weekend stuff, cache=TRUE}
#
# create weekend and weekday imputed step stuff. Loop through all the data and
# fill in the weekend and weekday vales.
#
for(i in 1:nrow(activity))
{
  if(activity$dweekend[i] == 1)
  {
    # 
    #    Note that I want predicted values to be nonnegative. Which I do below.
    #
    activity$weekend_steps[i] <- activity$imputed_steps[i]
    activity$weekday_steps[i] <- NA
    activity$day_class_factor <- "Weekend"
  }
  if(activity$dweekend[i] == 0)
  {
    # 
    #    Note that I want predicted values to be nonnegative. Which I do below.
    #
    activity$weekday_steps[i] <- activity$imputed_steps[i]
    activity$weekend_steps[i] <- NA
    activity$day_class_factor <- "Weekday"
  }
}
summary(activity$weekday_steps)
#
#
#
weekend_activity <- subset (activity,dweekend==1)
str(weekend_activity)
summary(weekend_activity)
weekday_activity <- subset (activity,dweekend==0)
str(weekday_activity)
summary(weekday_activity)

```

Set up a weekend factor

```{R set up week_activity, cache=TRUE}
activity$weekend_factor <- as.factor(activity$dweekend)
```

This code is to be ignored. Left here as a template for me for future use 

```{R graph xy plot}
#xyplot(week_activity$imputed_steps ~ week_activity$interval | factor(as.factor(week_activity$dweekend),type="l", xlab = "Interval",ylab = "Steps"))
# xyplot(week_activity$imputed_steps ~ week_activity$interval , xlab = "Interval",ylab = "Steps")
```

Now deal with daily averages using aggregate.

```{R now deal with daily averages, cache=TRUE}
#
# First compute average by day
#
day_imputed_steps <- aggregate(imputed_steps,list(activity_day= activity_date),sum)
day_imputed_steps$daily_steps <- day_imputed_steps$x
str(day_imputed_steps)
summary(day_imputed_steps$daily_steps)
#
# now average intervals
#
interval_imputed_steps <- aggregate(imputed_steps,list(activity_interval= interval),mean)
interval_imputed_steps$interval_steps_imputed <- interval_imputed_steps$x
str(interval_imputed_steps)
summary(interval_imputed_steps)
#
# day_imputed_steps
```

Now show weekend and weekday separately

```{R weekend and weekday intervals, cache=TRUE}
interval_imputed_weekend_steps <- aggregate(activity$weekend_steps,list(activity_interval= interval),mean)
interval_imputed_weekend_steps <- interval_imputed_weekend_steps$x
str(interval_imputed_weekend_steps)
summary(interval_imputed_weekend_steps)
```

Now find the interval with the largest average number of (imputed) steps, and the corresponding number of steps. 

One notice. This is a resubmission. I noticed during peer grading that I had made a careless error in my original code. The maximum answer was correct, but my answer of"835" had the benefit of seeing the correct code. 

```{R maximum imputed steps, cache=TRUE}
#
#  find the maximum imputed steps
#
max_interval <- interval_imputed_steps[which.max(interval_imputed_steps$interval_steps_imputed),1]  
max_interval
# 
# maximum value of steps
#
max(interval_imputed_steps$interval_steps_imputed)
interval_imputed_steps[which.max(interval_imputed_steps$interval_steps_imputed),3]
```

## Now show some graphs


``` {R graphs, cache=TRUE}
## log of steps
#
#     I commented out the logs here
#
hist(day_imputed_steps$daily_steps , 
      breaks = 10,  
      main="Histogram of daily steps", 
     xlab = "average number of steps per day--with imputation",
     ) 
#
# time series by day
#
plot(as.Date(day_imputed_steps$activity_day),day_imputed_steps$daily_steps, type="l", xlab = "Activity Day",ylab = "Daily Steps")
# xyplot(as.Date(day_imputed_steps$activity_day),day_imputed_steps$daily_steps|day_imputed_steps$dweekend)
day_imputed_steps$day_of_week <-weekdays(as.Date(day_imputed_steps$activity_day))
# ggplot(day_imputed_steps)+geom_line(aes(x=activity_day,y=daily_steps))
#
#   Time series by 5-minute interval
#
table(day_imputed_steps$day_of_week)

dSunday<-as.numeric((day_imputed_steps$day_of_week=="Sunday"))
dSaturday<-as.numeric((day_imputed_steps$day_of_week=="Saturday"))
day_imputed_steps$dweekend<-as.numeric(dSunday+dSaturday)
table(day_imputed_steps$dweekend)
for(i in 1:nrow(day_imputed_steps))
{
  day_imputed_steps$day_class[i]<- "weekday"
  if(day_imputed_steps$dweekend[i] == 1)
  {
    # 
    #    Note that I want predicted values to be nonnegative. Which I do below.
    #
    day_imputed_steps$day_class[i]<- "weekend"
  }
}
day_imputed_steps$day_class<- as.factor(day_imputed_steps$day_class)
table(day_imputed_steps$day_class)
xyplot(day_imputed_steps$daily_steps ~ as.Date(day_imputed_steps$activity_day) | factor(day_imputed_steps$day_class),type="l", xlab = "Activity Day",ylab = "Daily Steps")

hist(interval_imputed_steps$interval_steps_imputed, 
      breaks = 10,  
      main="Histogram of interval steps, averaged overall days",
      xlab = "average number of steps per interval--with imputation",
      ) 
plot(interval_imputed_steps$activity_interval,interval_imputed_steps$interval_steps_imputed, type="l",xlab = "Activity Interval",ylab = "Interval Steps")
xyplot(interval_imputed_steps$interval_steps_imputed ~ interval_imputed_steps$activity_interval, type="l",xlab = "Activity Interval",ylab = "Interval Steps")
```

Now weekend and weekday intervals

```{R weekend code, cache=TRUE}
#
# now average intervals
#
weekend_interval_imputed_steps <- aggregate(weekend_activity$imputed_steps,list(weekend_activity_interval= weekend_activity$interval),mean)
weekend_interval_imputed_steps$interval_steps_imputed <- weekend_interval_imputed_steps$x
#
# remove x
#
weekend_interval_imputed_steps$x <- NULL
#
# descriptives on weekend 
#
str(weekend_interval_imputed_steps)
summary(weekend_interval_imputed_steps)
```

Now weekday code

```{R weekday code, cache=TRUE}
#
# now average intervals
#
weekday_interval_imputed_steps <- aggregate(weekday_activity$imputed_steps,list(weekday_activity_interval= weekday_activity$interval),mean)
weekday_interval_imputed_steps$interval_steps_imputed <- weekday_interval_imputed_steps$x
str(weekday_interval_imputed_steps)
summary(weekday_interval_imputed_steps)
```

## Now show the weeken/weekday graphs and the multi-panel

```{R weekday and weekend graphs, cache=TRUE}
p1 <- plot(weekday_interval_imputed_steps$weekday_activity_interval,weekday_interval_imputed_steps$interval_steps_imputed, type="l",xlab = "Weekday Activity Interval",ylab = "Interval Steps")
p2 <- plot(weekend_interval_imputed_steps$weekend_activity_interval,weekend_interval_imputed_steps$interval_steps_imputed, type="l",xlab = "Weekend Activity Interval",ylab = "Interval Steps")

q1 <- qplot(weekday_interval_imputed_steps$weekday_activity_interval,weekday_interval_imputed_steps$interval_steps_imputed,xlab = "Weekday Activity Interval",ylab = "Interval Steps", geom=c("point", "line"))
q2 <- qplot(weekend_interval_imputed_steps$weekend_activity_interval,weekend_interval_imputed_steps$interval_steps_imputed,xlab = "Weekend Activity Interval",ylab = "Interval Steps", geom=c("point", "line"))

grid.arrange (q1, q2, nrow = 1)
```
